---
title: "Causal Regularization with and without Known Environments"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Causal Regularization with and without Known Environments}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6.5,
  fig.height = 4,
  fig.align = "center"
)
set.seed(1)
have_ggplot <- requireNamespace("ggplot2", quietly = TRUE)
```

## 1. Overview

Standard regression chooses the predictor that minimises average loss on the
data it sees. When the training data pools together several *environments*
(experimental batches, sites, time periods, sub-populations), that strategy can
latch onto associations that hold *on average* but are not stable across
environments — typically associations with **descendants of the response**
(spurious children) whose coupling to the outcome shifts from one environment to
another. A predictor that leans on such a feature looks great in-sample and
fails under distribution shift.

**Causal regularization** corrects for this by penalising predictors whose risk
is *unstable* across environments, steering the fit toward the stable (causal)
features. The `cr` package implements this idea in two modes that share the same
machinery and differ only in how the environments are obtained:

- **`cr(..., env = <labels>)` — Causal Regularization (CR), environments known.**
  You supply a two-level environment label and the method fixes the soft
  environment weights directly from it.
- **`cr(..., env = NULL)` / `acr(...)` — Adversarial Causal Regularization (ACR),
  environments unknown.** No labels are needed. The method *discovers* a
  worst-case split of the data into two pseudo-environments (Phase 1) and then
  regularizes against it.

In both modes a cross-validation step (Phase 2) selects the regularization
strength and solves the final estimator once.

```{r load}
library(cr)
```

## 2. A worked example: a spurious child

We simulate a small linear structural model. The response `Y` is *caused* by two
parents `X1` and `X2`. We then add a **spurious child** `X3` that is generated
*from* `Y` (so `X3` is a descendant, not a cause), and whose noise level changes
between two latent environments. Because `X3` tracks `Y` very tightly in
environment 0 and loosely in environment 1, its usefulness as a predictor is
*unstable* — exactly the kind of feature causal regularization should down-weight.

```{r sim}
set.seed(3)
n   <- 800
grp <- rep(c(0, 1), each = n / 2)        # latent environment (held out from ACR)

X1 <- rnorm(n)
X2 <- rnorm(n)
Y  <- 1.0 * X1 + 1.0 * X2 + rnorm(n, 0, 0.7)   # truth: X1 = X2 = 1, X3 = 0

# spurious child of Y, with environment-dependent noise -> unstable coupling
sd_env <- ifelse(grp == 0, 0.2, 1.5)
X3 <- Y + rnorm(n, 0, sd_env)

dat  <- data.frame(X1, X2, X3, Y)
envf <- factor(grp, labels = c("env0", "env1"))
```

Ordinary least squares cannot tell a parent from a child; it simply uses
whatever lowers the pooled loss. Here it over-weights the spurious child `X3`
and, as a consequence, *under-estimates* the true parents `X1` and `X2`:

```{r ols}
ols <- lm(Y ~ X1 + X2 + X3, data = dat)
round(coef(ols), 3)
```

The truth is `X1 = 1`, `X2 = 1`, `X3 = 0`, yet OLS puts a sizeable coefficient
on `X3` and pulls the parents below 1.

## 3. Phase 1, explained clearly

Phase 1 is the heart of the *unknown-environment* (ACR) mode. Its job is to
**discover** a split of the data into two pseudo-environments that makes the
predictor look as unstable as possible — without ever being told the true
environment labels. It works as follows.

**Step 1 — fit a pooled predictor once.** We fit an ordinary (ridge) least-squares
predictor on *all* the data and record its per-observation losses
$\ell_i = r_i^2$ (squared residuals). This pooled fit is computed **once** and
then held **fixed** for the rest of Phase 1.

**Step 2 — initialise soft weights from the signed residuals.** Each observation
gets a soft weight $w_i \in (0,1)$ that says how much it belongs to
pseudo-environment 1 (versus environment 2, which gets weight $1 - w_i$). We
seed these weights from the *signed* residuals of the pooled fit,

$$
\tilde\ell_i = \frac{r_i - \mathrm{median}(r)}{\mathrm{sd}(r) + \varepsilon},
\qquad
w_i = 0.1 + 0.8\,\sigma(\tilde\ell_i),
$$

with $\sigma$ the logistic function. Points the pooled model *over-predicts* and
points it *under-predicts* are nudged toward opposite pseudo-environments — a
natural starting guess for an unstable split.

**Step 3 — sharpen the split by gradient ascent on the risk contrast.** Define the
(normalised) weighted risks of the two pseudo-environments using the **fixed**
pooled losses,

$$
R_1(w) = \frac{\sum_i w_i\,\ell_i}{\sum_i w_i},
\qquad
R_2(w) = \frac{\sum_i (1 - w_i)\,\ell_i}{\sum_i (1 - w_i)}.
$$

Phase 1 then moves the weights by gradient **ascent** to make the two risks as
*different* as possible — it maximises the **risk contrast** $|R_1(w) - R_2(w)|$.
The ascent gradient is $s\,\bigl(\nabla R_1 - \nabla R_2^{(\text{chain})}\bigr)$
with $s = \mathrm{sign}(R_1 - R_2)$. Two regularizers keep the split sensible:

- a **balance penalty** $\kappa\,(\overline{w} - \rho)^2$ that keeps the mean
  weight near $\rho = 0.5$, so neither pseudo-environment collapses to a handful
  of points; and
- an **entropy** term that discourages the weights from saturating at exactly
  $0$ or $1$ too early.

> **Why does a high-contrast split expose causal instability?**
> A feature that is genuinely causal lowers the loss *the same way everywhere*,
> so no split of the data can make its associated risk swing much. A spurious,
> anti-causal feature, by contrast, helps a lot in some parts of the data and
> hurts in others. The split that *maximises* the risk gap therefore lines up
> with the dimension along which the predictor is unstable — which is precisely
> the spurious structure we want to regularize away. By searching for the
> worst-case split, ACR manufactures the "environments" that a labelled method
> would have been handed for free.

**No alternating re-fit.** This is the key structural point: Phase 1 **only moves
the weights**. The pooled predictor and its losses are computed once and never
re-solved inside the loop. (The predictor is re-estimated only afterwards, in
Phase 2, for the discovered weights.) This non-alternating form matches the
validated reference implementation and makes Phase 1 a clean, well-posed
maximisation over $w$.

Let us run ACR on the data and keep the Phase-1 diagnostics for plotting:

```{r fit-acr}
set.seed(3)
fit_acr <- acr(Y ~ X1 + X2 + X3, data = dat,
               control = cr_control(n_steps = 400))
round(coef(fit_acr), 3)
```

ACR shrinks the spurious child `X3` toward zero and pushes the parents back up
toward their true value of 1.

## 4. Phase 2, briefly

With the soft weights `w` in hand (discovered in ACR mode, or fixed from the
labels in CR mode), Phase 2:

1. builds the two **weighted moment matrices** for the pseudo-environments and
   their sum $G_+$ and difference $G_\Delta$ (likewise for the cross-moment
   vectors $Z_+, Z_\Delta$);
2. forms the regularized normal equations
   $(G_+ + \Gamma\, s\, G_\Delta + \lambda I)\,\beta = Z_+ + \Gamma\, s\, Z_\Delta$,
   where $s = \mathrm{sign}(R_1 - R_2)$ aligns the penalty with the heavier
   environment and the ridge $\lambda$ keeps the system PSD / well-conditioned;
3. selects the regularization strength $\Gamma$ by **K-fold cross-validation** on
   a worst-case risk criterion $(R_1 + R_2) + \Gamma\,|R_1 - R_2|$; and
4. **solves the estimator once** at the selected $\Gamma$.

```{r gamma}
fit_acr$gamma_fit   # selected regularization strength
```

## 5. Plots

```{r ggplot-theme, eval = have_ggplot}
library(ggplot2)
theme_set(theme_minimal(base_size = 12))
```

### Plot A — Phase-1 convergence

The Phase-1 history records the two pseudo-environment risks at every step. As
the split sharpens, $R_1$ and $R_2$ pull apart and the contrast $|R_1 - R_2|$
grows.

```{r plotA, eval = have_ggplot}
h <- fit_acr$history
conv <- data.frame(
  iteration = rep(seq_along(h$risk1), 3),
  value     = c(h$risk1, h$risk2, abs(h$risk1 - h$risk2)),
  series    = factor(rep(c("R1", "R2", "|R1 - R2|"),
                         each = length(h$risk1)),
                     levels = c("R1", "R2", "|R1 - R2|"))
)

ggplot(conv, aes(iteration, value, colour = series)) +
  geom_line(linewidth = 0.8) +
  scale_colour_manual(values = c("R1" = "#1b9e77",
                                 "R2" = "#d95f02",
                                 "|R1 - R2|" = "#7570b3")) +
  labs(title = "Phase 1: the split sharpens over iterations",
       x = "Iteration", y = "Pseudo-environment risk",
       colour = NULL) +
  theme(legend.position = "top")
```

### Plot B — discovered weights vs. the true environment

Phase 1 never sees `grp`. If it has recovered the latent split, the learned
weights should separate by the *true* (held-out) environment label.

```{r plotB, eval = have_ggplot}
wdf <- data.frame(weight = fit_acr$weights, environment = envf)

ggplot(wdf, aes(environment, weight, colour = environment)) +
  geom_jitter(width = 0.15, height = 0, alpha = 0.35, size = 1) +
  geom_boxplot(width = 0.4, outlier.shape = NA, alpha = 0,
               colour = "grey20") +
  scale_colour_manual(values = c("env0" = "#1b9e77", "env1" = "#d95f02")) +
  labs(title = "Discovered soft weights group by the true environment",
       x = "True (held-out) environment", y = "Learned weight w") +
  theme(legend.position = "none")
```

### Plot C — coefficient recovery

Finally we compare the estimated coefficients of OLS, ACR (unknown env), CR
(known env), and the truth. The spurious child `X3` should be shrunk toward zero
and the parents `X1`, `X2` recovered toward 1.

```{r fit-cr}
fit_cr <- cr(Y ~ X1 + X2 + X3, data = dat, env = envf)
round(coef(fit_cr), 3)
```

```{r plotC, eval = have_ggplot}
truth <- c(`(Intercept)` = 0, X1 = 1, X2 = 1, X3 = 0)
terms <- c("X1", "X2", "X3")

coef_df <- rbind(
  data.frame(term = terms, value = coef(ols)[terms],     method = "OLS"),
  data.frame(term = terms, value = coef(fit_acr)[terms], method = "ACR (env unknown)"),
  data.frame(term = terms, value = coef(fit_cr)[terms],  method = "CR (env known)"),
  data.frame(term = terms, value = truth[terms],         method = "Truth")
)
coef_df$method <- factor(coef_df$method,
                         levels = c("Truth", "OLS", "ACR (env unknown)", "CR (env known)"))

ggplot(coef_df, aes(term, value, fill = method)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_hline(yintercept = 0, colour = "grey60") +
  scale_fill_manual(values = c("Truth" = "grey30",
                               "OLS" = "#d95f02",
                               "ACR (env unknown)" = "#1b9e77",
                               "CR (env known)" = "#7570b3")) +
  labs(title = "Coefficient recovery: the spurious child X3 is shrunk",
       x = NULL, y = "Estimated coefficient", fill = NULL) +
  theme(legend.position = "top")
```

The two causal-regularization fits keep the parents `X1`, `X2` near 1 and pull
the spurious child `X3` toward 0, while OLS over-weights `X3`.

## 6. Known vs. unknown environments

The same data illustrates the two modes side by side:

```{r compare}
tab <- rbind(
  Truth = c(X1 = 1, X2 = 1, X3 = 0),
  OLS   = coef(ols)[c("X1", "X2", "X3")],
  ACR   = coef(fit_acr)[c("X1", "X2", "X3")],
  CR    = coef(fit_cr)[c("X1", "X2", "X3")]
)
round(tab, 3)
```

- **`acr(...)` (env unknown)** has to *discover* the worst-case split in Phase 1
  before it can regularize. It works from the data alone and recovers most of
  the structure.
- **`cr(..., env = envf)` (env known)** skips Phase 1 entirely and fixes the
  weights from the labels. With the true split handed to it, it regularizes
  against exactly the right instability and typically suppresses the spurious
  child most strongly.

Both share Phase 2 unchanged. In practice you use `cr(env = ...)` when reliable
environment labels exist, and `acr(...)` when they do not.
